---
title: 'Reproducible Research: Peeer Assessment 1'
author: "Vitor Zamprogno amancio Pereira"
date: "11/5/2017"
output: 
  html_document:
    keep_md: true
---

This R Markdown document is the results of Reproducible Research Peer-graded Assignment: Course Project 1
submitted by Vitor Zamprogno Amancio Pereiras


This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip


The variables included in this dataset are:

-**steps**: Number of steps taking in a 5-minute interval (missing values are coded as ð ™½ð ™°)
-**date**: The date on which the measurement was taken in YYYY-MM-DD format
-**interval**: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

### Initial settings
```{r}
rm(list=ls()) 
setwd("~/Documents/DataAnalyst/CouseraCourse/Reproducible-research/RepData_PeerAssessment1")
require(ggplot2)
require(dplyr)
require(lubridate)
```

## Loading and preprocessing the data

```{r}
if (!file.exists("activity.zip")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip","activity.zip", method="curl")
}
if (!file.exists("activity.csv")) {
  unzip("activity.zip")
}
activityFull=read.csv("activity.csv")
paste("Dataset with ",nrow(activityFull)," rows")  
head(activityFull)
```
We can see that the dataset contains lines with NA, we will then initially remove them
```{r}
activity = na.omit(activityFull)
head(activity)
```
## What is mean total number of steps taken per day?

#####Calculating total steps by date
```{r}
sumStepsActivity = activity %>% group_by(date) %>% summarise(totalSteps = sum(steps))
head(sumStepsActivity)
```

#####Let's do the Histogram of the total number of steps taken each day
```{r}
colores=c(rbind(colors()[grepl("orange",colors())][1:15],colors()[grepl("blue",colors())][1:15]))
g=ggplot(sumStepsActivity,aes(totalSteps)) + geom_histogram(fill=colores) 
g = g + labs(x="Number of steps taken in a day", y="Number of Occurrences", 
             title="Total number of steps taken each day" )
g = g + theme_get() 
print(g)
```

#####Mean and median number of steps taken each day
```{r}
paste("Mean of steps taken each day",mean(sumStepsActivity$totalSteps), sep="=")
paste("Median number of steps taken each day",median(sumStepsActivity$totalSteps), sep="=")
```


## What is the average daily activity pattern?

#####Calculating mean steps by interval
```{r}
IntervalActivity = activity %>% group_by(interval) %>% summarise(meanSteps = mean(steps))
head(IntervalActivity)
```

####Ploting
```{r}
g=ggplot(IntervalActivity,aes(interval,meanSteps)) + geom_line(colour="blue") 
g = g + labs(x="Average steps", y="intervals (5 min)", 
             title="Daily activity" )
g = g + theme_get() 
print(g)
```

####The 5-minute interval that, on average, contains the maximum number of steps
```{r}
paste("The 5-minute interval that contains maximum average of steps",
      IntervalActivity$interval[IntervalActivity$meanSteps==max(IntervalActivity$meanSteps)],
      sep="=")
paste("Average of steps at this interval",max(IntervalActivity$meanSteps),sep="=") 
```

## Imputing missing values

Initially (in previous steps) we simply removed the lines containing NA 
**"activity = na.omit (activityFull)"**

Now we will use a strategy to impute values by replacing NAs

####Analyzing the occurrence of NA in the dataset

######Which fields contains NA values?
```{r}
for(i in 1:ncol(activityFull)) {
#  label=paste("NA values in column ",names(activityFull)[i])
  print(paste(paste("NA values in column ",names(activityFull)[i]),sum(is.na(activityFull[,i]))))
}
```
**Only the column "steps" contains NA Values**


######How many and what percentage?
```{r}
paste("Total lines with NA",nrow(activityFull)-nrow(na.omit(activityFull)),sep="=")  
print(paste("Percentage of NA values",round(((nrow(activityFull)-nrow(na.omit(activityFull)))/nrow(activityFull))*100),"%"))
```

####The strategy adopted was to fill the NA with the average steps of the interval to which the record belongs.
####Will be created a new dataset with the NA filled in this way
```{r}
activityImp = merge(activityFull,IntervalActivity, by="interval") 
activityImp$steps[is.na(activityImp$steps)] = activityImp$meanSteps[is.na(activityImp$steps)]
head(activityImp)
activityImp =  activityImp[,1:3]
head(activityImp)
```


####Histogram of the total number of steps taken each day after missing values are imputed
```{r}
sumStepsActivityImp = activityImp  %>% group_by(date) %>% summarise(totalSteps = sum(steps))
head(sumStepsActivityImp)

colores=c(rbind(colors()[grepl("orange",colors())][1:15],colors()[grepl("blue",colors())][1:15]))
g=ggplot(sumStepsActivityImp,aes(totalSteps)) + geom_histogram(fill=colores) 
g = g + labs(x="Number of steps taken in a day", y="Number of Occurrences", 
             title="Total number of steps taken each day")
g = g + theme_get() 

print(g)

```

## Are there differences in activity patterns between weekdays and weekends?

####Creating a dataset containing the information whether the date is weekend or not ("weekend" or "weekday")
```{r}
activityKD = activityImp %>% 
             mutate(kindofday=ifelse(wday(as.Date(date, format = "%Y-%m-%d"))==1 | 
                                     wday(as.Date(date, format = "%Y-%m-%d"))== 7,"weekend","weekday"))
head(activityKD)
```

####Summarizing
```{r}
IntervalActivityKD = activityKD  %>% group_by(interval,kindofday) %>% summarise(meanSteps = mean(steps))
head(IntervalActivityKD)
```

####PLoting
```{r}
g=ggplot(IntervalActivityKD,aes(interval,meanSteps)) + geom_line(aes(colour=kindofday))
g = g + labs(x="Average steps", y="intervals (5 min)", 
             title="Daily activity (Weekends x Weekdays)" )
g = g + theme_get() + facet_grid(kindofday ~ .)
print(g)
```